---
title: "Evaluating an ensemble model of phonological categorisation on three variable morphological patterns in Hungarian"
author: "Rácz, Péter"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
# sed -i 's/Readme/ApocalypticLog/g' *.Rmd

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, error = FALSE, message = FALSE, fig.path = 'figures/', fig.width = 8, fig.height = 8)

# set wd in md
knitr::opts_knit$set(root.dir = '~/Github/Racz2024b')
setwd('~/Github/Racz2024b')

set.seed(1337)
options(knitr.kable.NA = "")

library(tidyverse)
library(glue)
library(magrittr)
library(knitr)
library(ggthemes)
library(lme4)
library(patchwork)
library(ggrepel)
library(stringdist)

# some of this takes a lot of iterations
no_cores <- future::availableCores() - 1
future::plan(future::multisession, workers = no_cores)

# source helper functions
source('code/helper.R')

# other functions

# -- fun -- #

# build 3d mds from dist table
makeDistMDS = function(dat){
  dat2 = dat |>
    dplyr::filter(segment1 != ' ', segment2 != ' ')
  dat_matrix = dat2 |>
    tidyr::pivot_wider(names_from = segment2, values_from = dist) |> 
    dplyr::select(-segment1) |> 
    as.matrix()
  dat_mds = stats::cmdscale(dat_matrix, k = 3)
  tidyr::tibble(
    x = dat_mds[,1],
    y = dat_mds[,2],
    z = dat_mds[,3],
    label = unique(dat2$segment1)
  ) 
}

# wrapper for KNN that filters for variation
wrapKNN = function(dat, distance_type, my_variation, var_p, var_s, var_k){
dat |>
  dplyr::filter(variation == my_variation) |>
  KNN(distance_type = distance_type, var_s = var_s, var_k = var_k, var_p = var_p)
}
# wrapper for GCM that filters for variation
wrapGCM = function(dat, distance_type, my_variation, var_s, var_p){
dat |>
  dplyr::filter(variation == my_variation) |>
  GCM(distance_type = distance_type, var_s = var_s, var_p = var_p)
}

# combine result tibbles with test responses
combineWithTest = function(dat){
  test %>% 
  rename(test = string) %>% 
  select(test,resp1,resp2,log_odds) %>% 
  inner_join(dat, by = 'test')
}

# fit a model that predicts response rations from category labels and pulls the summary
testAccuracy = function(dat){
  fit = glm(cbind(resp1,resp2) ~ 1 + category_high, data = dat, family = binomial(link = 'logit'))
  broom::tidy(fit) %>% 
    filter(term == 'category_high') %>% 
    mutate(sig = case_when(
      p.value < .001 ~ '***',
      p.value < .01 ~ '**',
      p.value < .05 ~ '*',
      p.value >= .05 ~ 'ns.'
      )
    )
}
```

In this readme, we go through a way of measuring distance between words based on natural classes that are, in turn, based on segmental similarity. Then we use this and other distances in three categorisation models to predict what people do in a Wug task with three variable patterns of Hungarian morphology.

There are three patterns of variation (see References for details):

- verb variation between `-k` and `-m` in the 1sg.indef of verbs
- verb variation between `CC.V+suffix` and `CVC.suffix` in a range of verbs, here focussing on the pl.indef
- noun variation between front vowel and back vowel suffixes in bisyllabic noun stems with a semi-transparent second vowel (`[e]` or `[é]`)

There are three models:

- K Nearest Neighbours
- the Generalised Context Model
- the Tiny Overlap Finder, an R-based algorithm under development, extremely heavily influenced by the Minimal Generalisation Learner. (I don't call this RMGL or something like that because the models do not give the same output)

There are three word distance measures:

- Levenshtein distance / edit distance
- Phonological distance, based on segmental distance across aligned word forms
- Jaccard distance

The Tiny Overlap Finder uses edit distance between strings and generally works very differently from the Generalised Context Model and the K-nearest Neighbours Model.

## Table of contents

In this readme, we will:

1. Generate natural classes from a language's phonological feature matrix
2. Generate segment-distances from the natural classes
3. Generate word-distances from the segment-distances using a word aligner
4. Generate categories from the word distances and an algorithm in the training data and apply these to test data
5. Generate generalised overlaps/rules from the words in the training data and see how these rules predict the test data

```{r preload}
# features
h = read_tsv('dat/feature_matrices/siptar_torkenczy_toth_racz_hungarian.tsv')
# nat classes
nch = read_tsv('dat/natural_classes/siptar_torkenczy_toth_racz_hungarian_nc.tsv')
# segment distance lookup
lookup_h = read_tsv('dat/segmental_distances/siptar_torkenczy_toth_racz_hungarian_dt.tsv')
# training and test data
test = read_tsv('dat/training_sets/test_set.tsv')
training = read_tsv('dat/training_sets/training_set.tsv')
# alignments of training x test data across three variations
# alignments_lakok = read_tsv('dat/alignments/alignments_lakok.tsv')
# alignments_cselekszenek = read_tsv('dat/alignments/alignments_cselekszenek.tsv')
# alignments_hotelban = read_tsv('dat/alignments/alignments_hotelban.tsv')
word_distance = read_tsv('dat/alignments/word_distances.tsv')

# tiny overlap finder things

# training files
training_tof = read_tsv('dat/training_sets/training_mgl.tsv')
# test files
test_tof = read_tsv('dat/training_sets/test_mgl.tsv')
# best rule sets
tof = read_tsv('dat/tof/best_tofs.tsv')
# predictions of said best rule sets
tof_pred = read_tsv('dat/tof/best_tof_predictions.tsv')
# models on tof predictions
load('dat/tof/tof_results.rda')
# illustrations for uh illustrative purposes
i_lakok_score = read_tsv('dat/tof/i_lakok_score.tsv')
i_lakok_pred = read_tsv('dat/tof/i_lakok_pred.tsv')

# this has to be done each time this is loaded
lookup_h %<>% mutate(across(-dist, ~ ifelse(is.na(.x), " ", .x)))
```

## Quick start

We do this below but in detail:

```{r quick_start, echo = T, eval = F}
# a. generate natural classes

nch = h |>
  generateNaturalClasses()

# b. build distance table for pairwise segment comparisons

lookup_h = buildDistTable(h, nch) |>
  addLevenshtein()

# c. align test and target words to find best phon-based alignment, here, for the 'lakok/lakom' variation. this takes ages.

alignments_lakok = runLookup(test_l,training_l,lookup_h)

# d. get distance based on best alignment, here, for lakok. join back with training data.

word_distance_lakok = alignments_lakok |>
  dplyr::distinct(test,training,total_dist) |>
  dplyr::rename(phon_dist = total_dist) |>
  dplyr::left_join(training, join_by('training' == 'string'))

# e. use the paired data to fit some sort of a phon distance based learning model, here, a KNN

# you can use a wrapper function to do a-e:
KNNwrapper(test = test, training = training, feature_matrix = fm, my_distance = 'phon', my_s = .1, my_k = 3, my_p = 1)

# you can use the wrapper function to use some other distance and skip the whole alignment bit:
KNNwrapper(test = test, training = training, feature_matrix = fm, my_distance = 'jaccard', my_s = .1, my_k = 3, my_p = 1)

# you can also use the wrapper function to use the TOF (see code/run_tof.r):
tof(training, alpha_upper = .75, alpha_lower = .75)
# this gives you rules and then you can use them to predict the test data

```

## 1. Generate natural classes

The feature matrix was adapted from Siptár and Törkenczy (2000). "ṯ", "ḏ", and "ṉ" are the palatal stops. "c" is "ts". Here is a sample of the feature matrix.

```{r feature_matrix}
h %>% 
  sample_n(10) %>% 
  # replace 0 with -
  mutate(across(-segment, ~ ifelse(.x == 0, '-', .x))) %>%
  # replace 1 with +
  mutate(across(-segment, ~ ifelse(.x == 1, '+', .x))) %>%
  # replace NA with " "
  mutate(across(-segment, ~ ifelse(is.na(.x), " ", .x))) %>%
  kable(caption = 'Table: Sample of the phonological feature matrix')
```

We turn it into underspecified natural class descriptions following Albrighty and Hayes (2003) and Frisch, Pierrehumbert, and Broe (2004).

```{r natural_classes_1, echo = T, eval = F}
nch = h |>
  generateNaturalClasses()

write_tsv(nch, 'dat/natural_classes/siptar_torkenczy_toth_racz_hungarian_nc.tsv')
```

Here's a sample of what the output looks like.

```{r natural_classes_2}
nch %>% 
  sample_n(10) %>% 
  kable(caption = 'Table: Sample of the natural classes')
```

## 2. Generate segment-distances from the natural classes

We use the natural classes to calculate phonological distances between segments and also add edit distances for the sake of completeness: Any segment vs any other segment has a phonological distance of 0-1, any segment vs nothing has a phonological distance of 1.

```{r segment_distances_1, echo = T, eval = F}
lookup_h = buildDistTable(h, nch) |>
  addLevenshtein()

write_tsv(lookup_h, 'dat/segmental_distances/siptar_torkenczy_toth_racz_hungarian_dt.tsv')
```

Here's a sample of what the output looks like.

```{r segment_distances_2}
lookup_h %>% 
  # replace all NA with ' ' in all columns
  mutate(across(all_of(c("segment1","segment2")), ~ ifelse(is.na(.x), ' ', .x))) %>% 
  sample_n(10) %>% 
  kable(digits = 2, caption = 'Table: Sample of the segmental distances')
```

Distances between objects can be turned into a map of objects using multidimensional scaling. Here's what segmental distances look like in the output as a map, across 3 dimensions.

```{r segment_distances_3, fig.width = 8, fig.height = 8}
# some innocent fun
mds_h = makeDistMDS(lookup_h)
# spec theme
my_theme = theme(
    axis.text.x = ggplot2::element_blank(),
    axis.text.y = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    plot.title = ggplot2::element_blank()
  )
# plot mds_h with text label
p1 = mds_h |>
  ggplot(aes(x = x, y = y, label = label)) +
  geom_label_repel(max.overlaps = 11) +
  theme_few() +
  my_theme
p2 = mds_h |>
  ggplot(aes(x = y, y = z, label = label)) +
  geom_label_repel(max.overlaps = 11) +
  theme_few() +
  my_theme
p3 = mds_h |>
  ggplot(aes(x = x, y = z, label = label)) +
  geom_label_repel(max.overlaps = 11) +
  theme_few() +
  my_theme
(p1 + p2) / (plot_spacer() + p3) + plot_annotation(title = 'Figure: Segmental distances in 3 dimensions')

```

These three dimensions don't exactly match to place or manner of articulation, but we can recognise patterns. The top left panel splits segments into vowels, non-coronals, coronal stops, and coronal continuants. The top right panel shows velars, labials, and coronals, mixing in the vowels, which share place of articulation features with the consonants. The bottom right panel splits vowels from consonants, and shows consonants in a continuum from labial and dorsal to coronal.

## 3. Generate word-distances from the segment-distances using a word aligner

We want to find the alignment between word 1 and word 2 where the pairwise comparison of their segments yields the smallest total distance. We can skip segment slots. This is a good alignment between "asztrál" [astra:l] (astral) and "osztály" [osta:j] (class):

```
a s t r á l
o s t   á j 
```

This isn't:

```
a s t r á l
o s t á j 
```

For each word pair, we do every conceivable alignment and pick the one with the shortest total distance. This will be the pair with the best alignment. This process is resource intensive. To run it, we need the two words and the lookup table generated in step 2. The resulting table shows the two strings in their bestalignment (col 1 and 2), the distances between the segments (dist, this is 0 for identical segments and 1 for deletion / insertion) and the total phonological distance (which is the same value for the whole alignment).

```{r word_distances_1, echo = T}
alignWords('astrál', 'ostáj', lookup_h) %>% 
  kable(digits = 2, caption = 'Table: Sample alignment 1')
```

Some of this is less intuitive, because sometimes the best alignment is ceteris paribus not the one you'd expect, see:

```{r word_distances_2, echo = T}
alignWords('astalka', 'astrál', lookup_h) %>% 
  kable(digits = 2, caption = 'Table: Sample alignment 2')
```

Apparently comparing the final `ka$` to the final `ál$` is a better bargain than skipping 1 segment and gaining a penalty of 1 and then comparing `ka $` to ` ál$`.

This is based on Albright and Hayes (2003) and Dawdy-Hesterberg and Pierrehumbert (2014).

## 4. Generate categories from the word distances and an algorithm

I have some training and test data for three variable patterns in Hungarian. See the preprint for details.

a. inflectional: 1sg.indef variation: lakok/lakom ("I live")
b. inflectional: vowel deletion: cselekszenek/cselekednek ("they act")
c. declensional: vowel harmony: hotelban/hotelben ("in the hotel")

- training data come from Hungarian webcorpus 2.
- test data come from a wug task where people responded to prompts in a forced-choice format. about 30 people responded to each prompt. they came from a student pool. See the Rácz & Lukács preprint for details. (The preprint only talks about 1 and 2 right now, but it covers methods and ethics for 3 as well.)

Training data look like this:

```{r training_data_1}
# see code/training_preprocessing.r for details
training %>% 
  group_by(variation,category) %>% 
  sample_n(2) %>% 
  kable(digits = 2, caption = 'Table: Sample of the training data')
```

The category labels are pretty arbitrary.

For 'cselekszenek/cselekednek', verbs that end in `CCik$` are the "high" category and verbs that end in `CVCik$` are the "low" category. (This is slightly more complicated: some `CVCik$` verbs act like `CCik$` verbs with vowel-initial suffixes and these were put in the `CCik$` category.)

For 'hotelban/hotelben', the two variants are back or front vowel variants of the same suffix (e.g. "ban/ben", "in") -- but these vary across a range of suffixes, the suffixes themselves have different total frequencies, so I fit a mixed model predicting `cbind(freq1,freq2)` in the corpus with a stem and suffix intercept. I then extracted the stem random effects and split the distribution across the median. This is the "high" and "low" category. Nouns that prefer back suffixes are in the "high" category and nouns that prefer front suffixes are in the "low" category.

For 'lakok/lakom', there is one exponent that varies and you can count the two forms (lakok, lakom) per verb and calculate log odds, then split the distribution across the median.

Training data was restricted to exclude less frequent forms and very long forms. This is fairly arbitrary.

Test data look like this:

```{r test_data}
# see code/training_preprocessing.r for details
test %>% 
  group_by(variation) %>% 
  select(base,variation,variant1,variant2,resp1,resp2,string) %>% 
  sample_n(5) %>% 
  kable(digits = 2, caption = 'Table: Sample of the test data')

```

Test data were generated from corpus distributions using a syllabic constituent-based ngram model. The resulting words are somewhat onset-heavy (words in Hungarian can begin with `^sp` but maybe not this often). Note that for 'cselekszenek', the Wug prompts had a CC and CVC form, as both are possible (that being the point). We use the CC form to calculate distances. This is a semi-arbitrary decision: variable verbs have a CC and a CVC form. Stable CVC verbs never have a CC form. See the preprint for details.

`string` is transcribed using a simple converter that maps Hungarian orthography to IPA.

For each variation, you pair test and training words and calc word distance. For the verbs ('cselekszenek', 'lakok' types) you can drop the `ik$` ending as all verbs have it anyway.

```{r word_distances_3, echo = T, eval = F}
# this legitimately takes a while.
# for stand-alone code go to code/run_alignment.r

# set up test and training data
testl = filter(test, variation == 'lakok/lakom')
trainingl = filter(training, variation == 'lakok/lakom')
testcs = filter(test, variation == 'cselekszenek/cselekednek')
trainingcs = filter(training, variation == 'cselekszenek/cselekednek')
testh = filter(test, variation == 'hotelban/hotelben')
trainingh = filter(training, variation == 'hotelban/hotelben')

# run the lookup for all variations
alignments_lakok = runLookup(testl,trainingl,lookup_h)
alignments_cselekszenek = runLookup(testcs,trainingcs,lookup_h)
alignments_hotelban = runLookup(testh,trainingh,lookup_h)

# write_tsv(alignments_lakok, 'dat/alignments/alignments_lakok.tsv')
# write_tsv(alignments_cselekszenek, 'dat/alignments/alignments_cselekszenek.tsv')
# write_tsv(alignments_hotelban, 'dat/alignments/alignments_hotelban.tsv')
```

We now have best alignments for training : test data. These might be useful someday, like if we want to build a rule-based learner. They look like this. In the test col are three nonce words used in the Wug task. In the training col are three aligned training words. Segment 1 and 2 show the test and training words broken up in segments, in their best alignment. Dist shows segmental distances (this is the same as the table above). Total dist shows the total distance for the alignment, which is the sum of segmental distances. 

This is just three examples. Note that we deleted the verbal "ik" ending because that's the same everywhere and this speeds up computation.

```{r word_distances_4, eval = F}

keep = alignments_lakok %>%
  distinct(test,training) %>% 
  sample_n(3)

alignments_lakok %>% 
  inner_join(keep) %>%
  # replace NA with ' '
  mutate(segment1 = ifelse(is.na(segment1), ' ', segment1),
         segment2 = ifelse(is.na(segment2), ' ', segment2)) %>%
  kable(digits = 2)

# the same in a md Table: 
```


|test   |training |segment1 |segment2 | dist|  phon_dist|
|:------|:--------|:--------|:--------|----:|----------:|
|narádz |bábáškod |n        |b        | 0.84|       5.84|
|narádz |bábáškod |a        |á        | 0.58|       5.84|
|narádz |bábáškod |r        |b        | 0.94|       5.84|
|narádz |bábáškod |á        |á        | 0.00|       5.84|
|narádz |bábáškod |         |š        | 1.00|       5.84|
|narádz |bábáškod |d        |k        | 0.84|       5.84|
|narádz |bábáškod |         |o        | 1.00|       5.84|
|narádz |bábáškod |z        |d        | 0.63|       5.84|
|šalárz |kérked   |š        |k        | 0.86|       4.21|
|šalárz |kérked   |a        |é        | 0.81|       4.21|
|šalárz |kérked   |l        |r        | 0.12|       4.21|
|šalárz |kérked   |á        |k        | 0.92|       4.21|
|šalárz |kérked   |r        |e        | 0.88|       4.21|
|šalárz |kérked   |z        |d        | 0.63|       4.21|
|streml |vesőd    |s        |v        | 0.87|       5.11|
|streml |vesőd    |t        |e        | 0.95|       5.11|
|streml |vesőd    |r        |s        | 0.64|       5.11|
|streml |vesőd    |e        |ő        | 0.79|       5.11|
|streml |vesőd    |m        |d        | 0.86|       5.11|
|streml |vesőd    |l        |         | 1.00|       5.11|

Table: Example alignments

What we actually need is the total phonological distances.

```{r add_distance, eval = F, echo = F}
alignments_lakok_2 = alignments_lakok %>% 
  mutate(variation = 'lakok/lakom') %>% 
  distinct(test,training,phon_dist,variation)

alignments_hotelban_2 = alignments_hotelban %>% 
  mutate(variation = 'hotelban/hotelben') %>% 
  distinct(test,training,phon_dist,variation)

alignments_cselekszenek_2 = alignments_cselekszenek %>% 
  mutate(variation = 'cselekszenek/cselekednek') %>%
  distinct(test,training,phon_dist,variation)

word_distance = bind_rows(
  alignments_lakok_2,
  alignments_hotelban_2,
  alignments_cselekszenek_2
)

write_tsv(word_distance, 'dat/alignments/word_distances.tsv')
```

The final result looks like this. We have total phonological distance, based on best alignment, between each test word and each training word in each variation pattern.

```{r word_distances_5}

word_distance %>% 
  group_by(variation) %>% 
  sample_n(5) %>%
  select(test,training,variation,phon_dist) %>% 
  kable(digits = 2, caption = 'Table: Word distances for nonce words in the Wug task and their aligned training words.')
  
```

## 4. Modelling

The nonce words got variable A or B responses in the Wug task and have a specific similarity to existing words in two categories: A and B. A and B are different for each word type ('cselekszenek', 'hotelban', 'lakok'). We want to see whether similarity to existing words predicts variable responses. 

Training categories are:

- cselekszenek/cselekednek: stable `CCik$` verbs are high, stable `CVCik$` verbs are low
- lakok/lakom: variable verbs above the median log odds of k/m in the corpus are high, others are low
- hotelban/hotelben: we aggregated over multiple suffixed variants to have a random intercept for each stem, those over the median of random intercepts are high, others low

Training sets were also slightly culled to include only more frequent verbs and shorter verbs. (see preprint and `training_preprocessor.R` for details).

```{r categorisation2, echo = F}
# same verb if variable might be in both categories, for cselekszenek. double check it please.

word_distance = training %>% 
  rename(training = string, training_label = base) %>%
  distinct(training,training_label,variation,category) %>% 
  inner_join(word_distance) %>% 
  distinct(test,training,training_label,variation,category,phon_dist)

word_distance = test %>%
  rename(test = string, test_label = base) %>%
  inner_join(word_distance)
```

The end result looks like this. This is distances between test words and training words and responses to test words and training word category.

```{r categorisation3}
word_distance %>% 
  group_by(variation,category) %>%
  sample_n(2) %>%
  select(test_label,training_label,category,phon_dist) %>% 
  kable(digits = 2, caption = 'Table: Word distances for nonce words in the Wug task and their aligned training words, with training word category and test word response log odds (large log odds: lot of "high" responses).')

```

### K-Nearest Neighbour

We categorise test words based on similarity with training words using a K-Nearest Neighbours algorithm. The algorithm calculates word distance using the formula `exp ( - dist / s ) ^ p`.

We try `k = [1,3,5,7,15]`, `p = [1,2]`, and `s = [.1,.3,.5,.7,.9]` and `dist = [edit distance, jaccard distance, phonological distance]`. Edit distance tallies up the number of edits needed to get from Word A to Word B. Jaccard distance is the set size of the intersect of segments in Word A and Word B divided by the set size of the union of segments in Word A and Word B. Phonological distance is the whole setup above.

A regular old KNN compares the target word to the nearest n words and counts category labels. So if this is 3 words and they are "high", "high", "low", the target word is categorised as "high". My KNN outputs the mean category label, which, in the example, would be .66.

For each test word, we get a KNN prediction for each parameter combination. Then we fit a GLM that predicts how people "voted" on the word (like, 15 picked variant A and 20 picked variant B) from the model's category weight. We then grab the term "category: high" from the model and take the est, std error, z value (statistic), and p value. These express how well the KNN weights predict whether more people picked variant A or B. We compare models using the z value (statistic), which favours the least noise over the largest effect.

```{r categorisation4, echo = T}

# set up parameters
my_parameters_1 = crossing(
  var_p = c(1,2),
  var_k = c(1,2,3,5,7,15),
  var_s = c(.1,.3,.5,.7,.9),
  distance_type = c('edit','jaccard','phon'),
  variation = c('lakok/lakom','hotelban/hotelben','cselekszenek/cselekednek')
)

# run 540 models, parallelised, add prediction tibbles to each model
my_knns = my_parameters_1 %>% 
  mutate(
    knn = furrr::future_pmap(list(variation,distance_type,var_p,var_s,var_k), ~ wrapKNN(
      dat = word_distance,
      my_variation = ..1,
      distance_type = ..2,
      var_p = ..3,
      var_s = ..4,
      var_k = ..5
      )
    )
  )

# combine with test and calc accuracy
my_knns %<>% 
  mutate(
    knn2 = map(knn,combineWithTest),
    accuracy = map(knn2,testAccuracy)
  )

my_knns2 = my_knns %>% 
  select(-knn,-knn2) %>% 
  unnest(accuracy)

```

```{r categorisation7}
# get best model for each var

my_knns3 = my_knns2 %>% 
  ungroup() %>% 
  group_by(variation) %>% 
  arrange(-statistic) %>% 
  slice(1)

my_knns3 %>% 
  select(-term,-p.value) %>% 
  kable(digits = 2, caption = 'Table: Best KNN model for each variation.')
```

All three models can explain some variation in the test data. The 'cselekszenek' model uses phonological distance and the largest k, k=15. For 'hotelban', the best model also uses the largest k and jaccard distance. For 'lakok', the best model uses k=7 and edit distance. All three best models have an s = .1 which relatively penalises more distant neighbours over closer neighbours.

We can look at the best model for each segment distance type and variation.

```{r categorisation8}
my_knns2 %>% 
  ungroup() %>% 
  group_by(variation,distance_type) %>% 
  # filter(sim == 'phon') %>% 
  arrange(-statistic) %>% 
  select(-p.value) %>% 
  slice(1) %>% 
  kable(digits = 2, caption = 'Table: Best KNN model for each variation and distance type.')

```

For 'cselekszenek' and 'hotelban', the models don't differ much in accuracy depending on distance. For 'lakok', jaccard distance is markedly more terrible than the others. A K nearest-neighbour model based on phonological distance between test and training words is nearly on par with one based on edit distance, despite the extra information available to the former: the edit distance between [f] and [p] is the same as between [f] and [u], even though the former pair share more natural classes and so have a smaller phonological distance.

### Generalised Context Model

We categorise test words based on similarity with training words using the Generalised Context Model. We try `p = [1,2]`, `s = [.1,.3,.5,.7,.9]` and edit distance, jaccard distance, and phonological distance.

```{r categorisation6gcm, echo = T}
# we do the same thing as with the KNNs. If I did this once more I'd need to put it into a function! cheeky

# set up parameters
my_parameters_2 = crossing(
  var_p = c(1,2),
  var_s = c(.1,.3,.5,.7,.9),
  distance_type = c('edit','jaccard','phon'),
  variation = c('lakok/lakom','hotelban/hotelben','cselekszenek/cselekednek')
)

# run models
my_gcms = my_parameters_2 %>% 
  mutate(
    gcm = furrr::future_pmap(list(distance_type,variation,var_p,var_s), ~ wrapGCM(
      dat = word_distance,
      distance_type = ..1,
      my_variation = ..2,
      var_s = ..3,
      var_p = ..4
    ))
  )

# combine with test and calc accuracy
my_gcms %<>% 
  mutate(
    gcm2 = map(gcm,combineWithTest),
    accuracy = map(gcm2,testAccuracy)
  )

my_gcms2 = my_gcms %>% 
  select(-gcm,-gcm2) %>% 
  unnest(accuracy)
```

```{r categorisation7gcm}
# get best model for each var

my_gcms3 = my_gcms2 %>% 
  ungroup() %>% 
  group_by(variation) %>% 
  arrange(-statistic) %>% 
  select(-p.value) %>% 
  slice(1)

my_gcms3 %>% 
  kable(digits = 2, caption = 'Table: Best GCM model for each variation.')
```

Again, the categorisation models account for some variation across test words in all three variable patterns. For 'cselekszenek', the best model uses edit distance and an exponential distance metric. For 'lakok', the best model uses phonological distance and the exponential metric. For 'hotelban' it, again, uses jaccard distance.

```{r categorisation8gcm}
my_gcms2 %>% 
  ungroup() %>% 
  group_by(variation,distance_type) %>% 
  # filter(sim == 'phon') %>% 
  arrange(-statistic) %>% 
  select(-p.value) %>% 
  slice(1) %>% 
  kable(digits = 2, caption = 'Table: Best GCM model for each variation and distance type.')

```

## 5. The Tiny Overlap Finder

The tiny overlap finder looks for overlapping alternations in the training data and tries to generalise these. It was written in R following the steps in Albright and Hayes (2003) and, where they refer to it, Mikheev (1997). The tiny overlap finder looks for input-output alternations of the format `A -> B` in context `C _ D` and then tries to generalise these.

|no |input      |output       |rule     |context |
|:--|:----------|:------------|:--------|:-------|
| 1 |abc__x__   |abc__y__     | x -> y  | abc _  |
| 2 |bbc__x__   |bbc__y__     | x -> y  | bbc _  |
| 3 |adc__x__   |adc__y__     | x -> y  | adc _  |
| 4 |aec__x__   |aec__z__     | x -> z  | aec _  |

For the input-output pairs in the table above, we observe `x` turning into `y` or `z` word-finally. We can consolidate examples 1-2 to rule (i): `x -> y / bc _` and example 1-3 to rule (ii): `x -> y / c _`. Note that example (4) matches the structural description of rule (ii) but the rule does not apply to it. So rule (i) has a scope of 2 examples and matches 2 of these, while rule (ii) has a scope of 4 and matches 3. Rule (i) is more _reliable_ than rule (ii) as it applies to more of its context examples but rule (ii) has more examples in its context altogether.

We can adjust the reliability of our rules based on how many forms they apply to. This will express the notion that rule (i) is more reliable but more limited than rule (ii). We can also adjust rule reliability based on whether some rules do the work of other rules. In our example, rule (ii) is redundant in examples 1-2 where rule (i) does all its work. We can reflect this by adjusting down how much we confide in rule (ii).

The Minimal Generalisation Learner can use phonological distance to create generalised contexts. The tiny overlap finder only looks for segment-to-segment matches. Hungarian verbs tend to have strong templatic effects with many derivational and pseudo-derivational endings. These can have predictive effects on inflectional variation. The Tiny Overlap Finder can find these.

---

The tiny overlap finder notably doesn't give the same results on the same data as the java implementation of the minimal generalisation learner, for reasons I couldn't ascertain using either the MGL code or documentation. Further work is needed to bring the two in line. The tiny overlap finder clearly works, so it is included in this discussion.

---

First, we hard-wire phonological variation in our dataset because we want the tiny overlap finder to focus on the morphology (-k/-m, CC/CVC, -front/-back) instead of the phonology. (The K-nearest neighbour model and the generalised context model only had to pick between two variants, and the tiny overlap finder shouldn't have a much harder time than these.)

Here are the transformed input-output pairs in the training sets:

```{r tof1}
training_tof %>% 
  group_by(variation,suffix,category) %>% 
  sample_n(1) %>% 
  select(-output1) %>% 
  kable(caption = 'Table: Training data, tiny overlap finder')

```

For 'lakok' and 'cselekszenek', vowel harmony is not relevant, so we remove the final stem vowel and the linking vowel and the suffix vowel and replace these with V. For 'hotelban', the suffix vowel is precisely the point, so we keep it.

We transform test data the same way. We assign categories the same way we did for the KNN and GCM training sets: if a test form does A a lot, we assign it to A. If it does B a lot, we assign it to B.

```{r tof2}
test_tof %>% 
  rename(output = output2) %>% 
  mutate(
    suffix = ifelse(variation == 'lakok/lakom', '3sg', suffix),
    category = ifelse(log_odds > mean(log_odds), 'high', 'low')
    ) %>% 
  group_by(variation,suffix,category) %>% 
  slice(1) %>% 
  select(base,variation,suffix,category,input,output) %>% 
  kable(caption = 'Table: Test data, tiny overlap finder')

```

We fit the model separately on every variable suffix type so it doesn't need to distinguish those either. There are three variable suffixes for 'cselekszenek' and 'hotelban' and only one for 'lakok'. As noted earlier, the tiny overlap finder has two parameters, one penalising rules that cover fewer versus more input-output pairs in the training (as in i-ii in the examples above) and one penalising rules that have subrules that do most of the work for them (also as in i versus ii in the examples above). We try [.25,.5,.75,.9] for both of these parameters (for details see code/run_tof.r and code/tof.r and code/test_tof.r). We fit the tiny overlap finder on the suffix sets and then combine all rules for each variation pattern. Here are ten random rules for 'lakok', with no guidance:

```{r tof3}
tof %>% 
  filter(variation == 'lakok/lakom') %>% 
  mutate(
    some_examples = stringr::word(examples,1,5) %>% str_replace(',$', ''),
    some_exceptions = stringr::word(exceptions,1,5) %>% str_replace(',$', '')
  ) %>% 
  select(rule,scope,hits,reliability,alpha_lower,lower_confidence_limit,alpha_upper,impugned_lower_confidence_limit,some_examples,some_exceptions) %>% 
  sample_n(5) %>% 
  kable(caption = 'Table: Some rules for lakok/lakom', digits = 2)
```

Each rule is in the "input ending -> output ending / context" format. Scope is the number of words in the training data that match the context of the rule. Hits is the number of words that actually obey the rule. Reliability is hits / scope. To put it very simply, the lower confidence limit is the lower limit of our confidence of how well the rule would work for many forms, adjusted by our alpha value for the lower confidence limit. Impugned lower confidence limit is one of two things: if, given an arbitrary alpha value for the upper confidence limit, the rule does substantially more work than its sub-rules, it is the rule's own lower confidence limit. If the sub-rules do too much work, it's the best sub-rules upper confidence limit.

Here are ten rules for 'cselekszenek':

```{r tof4}
tof %>% 
  filter(variation == 'cselekszenek/cselekednek') %>% 
  mutate(
    some_examples = stringr::word(examples,1,5) %>% str_replace(',$', ''),
    some_exceptions = stringr::word(exceptions,1,5) %>% str_replace(',$', '')
  ) %>% 
  select(rule,scope,hits,reliability,alpha_lower,lower_confidence_limit,alpha_upper,impugned_lower_confidence_limit,some_examples,some_exceptions) %>% 
  sample_n(5) %>% 
  kable(caption = 'Table: Some rules for lakok/lakom', digits = 2)
```

Here are ten rules for 'hotelban':

```{r tof5}
tof %>% 
  filter(variation == 'hotelban/hotelben') %>% 
  mutate(
    some_examples = stringr::word(examples,1,5) %>% str_replace(',$', ''),
    some_exceptions = stringr::word(exceptions,1,5) %>% str_replace(',$', '')
  ) %>% 
  select(rule,scope,hits,reliability,alpha_lower,lower_confidence_limit,alpha_upper,impugned_lower_confidence_limit,some_examples,some_exceptions) %>% 
  sample_n(5) %>% 
  kable(caption = 'Table: Some rules for lakok/lakom', digits = 2)
```

We choose the two parameters, alpha_upper and alpha_lower, using grid search. Models are evaluated the following way. First, we match the rules to the test words. For each test word, we have two outputs in the forced choice task. We match each of these to rules generated by the tiny overlap learner and find the best rule for each output, based on the impugned lower confidence limit, as seen in the table below. 

```{r tof6}
i_lakok_pred %>% 
  arrange(base) %>% 
  arrange(base,variant,-impugned_lower_confidence_limit) %>% 
  mutate(
    some_examples = stringr::word(examples,1,5) %>% str_replace(',$', ''),
    some_exceptions = stringr::word(exceptions,1,5) %>% str_replace(',$', '')
  ) %>% 
  select(base,input,variant,output,rule,impugned_lower_confidence_limit,some_examples,some_exceptions) %>% 
  slice(1:10) %>% 
  kable(caption = 'Table: Some predictions for lakok/lakom', digits = 2)
```

We then take the best rule for each output and get an overall score by dividing the impugned lower confidence of the rule for output 1 with the sum of the impugned lower confidence of the rules for output 1 and 2. If the tiny overlap finder finds no rule for output 1 for a test word, the score will be 0 and vice versa. We can compare this score with the responses given by the participants.

```{r tof7}
i_lakok_score %>% 
  arrange(base) %>% 
  slice(1:10) %>% 
  select(base,resp1,resp2,log_odds,tof_score) %>% 
  kable(caption = 'Table: Some scores for lakok/lakom', digits = 2)
```

We use a generalised linear model predicting `cbind(resp1,resp2)` from the model score to find the best model for each variable pattern. The best model has the strongest predictor for score. This is the same as what we did for the K-Nearest Neighbours Model and the Generalised Context Model.

```{r tof8}
results %>% 
  mutate(
    sig = case_when(
      p.value < .001 ~ '***',
      p.value < .01 ~ '**',
      p.value < .05 ~ '*',
      p.value >= .05 ~ 'ns.'
      )
  ) %>% 
  select(-glm,-term,-p.value) %>% 
  kable(caption = 'Table: Best tiny overlap finder models for each variable pattern', digits = 2)
```

The tiny overlap finder is able to predict participant behaviour to some extent for all three patterns.

## The ensemble model

So what is the best at predicting what people will do with morphological variation in a forced-choice task using nonce forms? The K-Nearest Neighbours model, the Generalised Context Model, or the tiny overlap finder?

```{r best_models}
my_knns4 = my_knns2 %>% 
  select(distance_type,var_s,var_p,var_k,variation,estimate,std.error,statistic,sig) %>% 
  mutate(model = 'KNN')

my_gcms4 = my_gcms2 %>%
  select(distance_type,var_s,var_p,variation,estimate,std.error,statistic,sig) %>% 
  mutate(model = 'GCM')

my_tofs4 = results %>% 
  mutate(
    sig = case_when(
      p.value < .001 ~ '***',
      p.value < .01 ~ '**',
      p.value < .05 ~ '*',
      p.value >= .05 ~ 'ns.'
      ),
    model = 'TOF',
    distance_type = 'edit'
  ) %>% 
  select(model,distance_type,alpha_upper,alpha_lower,variation,estimate,std.error,statistic,sig) 

best_of_the_worst = bind_rows(my_gcms4,my_knns4,my_tofs4) %>% 
  select(variation,model,estimate,std.error,statistic,sig,distance_type,var_s,var_p,alpha_lower,alpha_upper) %>% 
  group_by(variation,model) %>%
  arrange(-statistic) %>%
  slice(1)

best_of_the_worst %>% 
  kable(digits = 2, caption = 'Table: Best best Generalised Context Model (GCM), K-nearest neighbours model (KNN), and tiny overlap finder (TOF) model for each variation.')
```

A lot is going on in this table. The table lists the best Generalised Context Model (GCM), K-nearest neighbours model (KNN), and tiny overlap finder (TOF) model for the three types of variation and also lists model parameters: distance type (this is always edit distance for the tiny overlap finder or TOF), s and p (only applicable to the distance calculations of the KNN and the GCM), and the alphas for the lower and upper confidence intervals (only applicable to the rule generalisations of the tiny overlap finder). The table also lists the estimates, standard errors, and statistics that express the predictive power of each model for participant responses.

For 'cselekszenek', the best model is the generalised context model, which uses similarity to all training forms across categories. This model calculates word distance using edit distance rather than aligned phonological dissimilarity. 

For 'hotelban', the best model is the generalised context model, using jaccard distance. 

For 'lakok', the best model is the tiny overlap finder, which by default uses segment-level similarity or edit distance to find overlapping contexts across training words.

An equally important question is whether the models individually contribute to explaining variation in the test data. To find out, we fit three generalised linear models, each predicting the odds of resp1 and resp2 in the three variable sets, using the best tiny overlap finder, generalised context model, and k-nearest neighbour model predictions together. We scale predictors and check for collinearity.

```{r best_models2}
tof_pred2 = tof_pred %>% 
  select(base,tof_score) %>% 
  rename(tof = tof_score)
  
knn_pred = my_knns %>% 
  inner_join(my_knns3) %>% 
  select(knn2) %>% 
  unnest(knn2) %>% 
  rename(knn = category_high, string = test) %>% 
  select(string, knn)

gcm_pred = my_gcms %>% 
  inner_join(my_gcms3) %>% 
  select(gcm2) %>% 
  unnest(gcm2) %>% 
  rename(gcm = category_high, string = test) %>% 
  select(string, gcm)

tests = test %>% 
  left_join(tof_pred2) %>% 
  left_join(knn_pred) %>% 
  left_join(gcm_pred) %>% 
  group_by(variation) %>% 
  mutate(
    tof = scales::rescale(tof),
    gcm = scales::rescale(gcm),
    knn = scales::rescale(knn),
  ) %>% 
  ungroup()

terms = tibble(terms = c(
  'tof + knn + gcm',
  'tof + gcm',
  'tof + knn',
  'gcm + knn'
  ))

tests2 = tests %>% 
  group_by(variation) %>% 
  nest() %>% 
  crossing(terms) %>% 
  mutate(
    formula = glue('cbind(resp1,resp2) ~ 1 + {terms}'),
    glm = map2(formula, data, ~ glm(formula = as.formula(.x), data = .y, family = binomial))
  ) %>% 
  select(variation,terms,glm) %>% 
  pivot_wider(names_from = terms, values_from = glm) %>% 
  mutate(
    gcm = map2(`tof + knn + gcm`, `tof + knn`, ~ performance::test_likelihoodratio(.x,.y)),
    knn = map2(`tof + knn + gcm`, `tof + gcm`, ~ performance::test_likelihoodratio(.x,.y)),
    tof = map2(`tof + knn + gcm`, `gcm + knn`, ~ performance::test_likelihoodratio(.x,.y))
  )

tests2 %>% 
  mutate(
    sum = map(`tof + knn + gcm`, ~ broom::tidy(.))
  ) %>% 
  select(variation,sum) %>% 
  unnest(sum) %>% 
  kable(digits = 2, caption = 'Ensemble model estimates for the three variable patterns.')

```

The term estimates for generalised linear models predicting test responses from the tiny overlap finder, k-nearest neighbour, and generalised context model scores across the three variable patterns are above. 

The k-nearest neighbour model does not explain model variation in any of the three models. The rules identified by the tiny overlap finder contribute to variation in the two verbal patterns, 'cselekszenek' and 'lakok'. For the latter, it is the most important predictor. For the noun pattern, 'hotelban', the only relevant predictor is the generalised context model.

We use a chi-square test of goodness of fit to check whether dropping each term from each model significantly reduces model fit. The results can be seen below.

```{r best_models3}

tests2 %>% 
  select(variation,gcm,knn,tof) %>% 
  pivot_longer(-variation) %>% 
  unnest(value) %>% 
  filter(Name == ".y") %>% 
  select(variation,name,Chi2,p) %>% 
  kable(digits = 2, caption = 'Likelihood tests for terms in the ensemble models.')

```

This test supports the results above. The k-nearest neighbour model has nothing to add. For the noun pattern, only the generalised context model is relevant.

We can calculate McFadden's pseudo R squared (1 - deviance / null deviance) for each model.

```{r best_models4}
# mcfadden's r squared
tests2 %>% 
  mutate(
    `mcfadden's r` = map(`tof + knn + gcm`, ~ with(summary(.), 1 - deviance/null.deviance))
  ) %>% 
  select(variation,`mcfadden's r`) %>% 
  unnest(`mcfadden's r`) %>% 
  kable(digits = 2, caption = 'Table: McFadden\'s pseudo R squared for the GLMs with the joint predictions.')

```

The 'cselekszenek' pattern is hard. Many different words vary in many different ways in the lexicon and participants generalise this in a very noisy way. For 'hotelban', form-based similarity is a reasonable measure to differentiate words that prefer back or front vowel suffixes.

For 'lakok', we get a remarkably good result, for the cognitive sciences. Here, both overall similarity and the derivational and pseudo-derivational endings found by the tiny overlap finder are useful. It also helps that this pattern is restricted to one paradigm slot, 1sg.indef and varies across many verbs consistently.

## Visualisations

The best model for each model type and variation. Model scores are rescaled for each variation.

```{r viz, fig.width = 9, fig.height = 9}

tests %>% 
  mutate(base2 = str_extract(base, '^[^ ]+(?=( |$))')) %>%
  select(variation,base2,log_odds,tof,knn,gcm) %>% 
  pivot_longer(cols = c(tof,knn,gcm), names_to = 'model', values_to = 'weight') %>% 
  ggplot(aes(x = weight, y = log_odds)) +
  geom_point() +
  theme_bw() +
  geom_smooth(method = mgcv::gam) +
  facet_wrap(~ variation + model, scales = 'free', nrow = 3)
```

## Discussion

People very clearly rely on the ambient language in making their wug task judgements on these three variable patterns. We can tell because the decisions they make for test words are partly based on the corpus distributions of existing words that the test words look like. The corpus is our best guess on the distributions in the ambient language. This is all of course in line with the existing literature.

It is not the case that people compare training words to one or two nearest neighbours and make their choices based on that. The best KNN models use far more neighbours and even these are beaten by more general models. Either all participants in the sample have a different nearest neighbour in mind for each test word, which seems unlikely, or they aggregate over a larger similarity space.

Phonological similarity is useful in the models, but it does not necessarily beat other measures of similarity when calculating how much a given test word looks like our training categories. 

For verbs, the tiny overlap finder is useful. This is because the verbal inflection patterns are sensitive to derivational and pseudo-derivational endings. The overlap finder can find these.

## References

The Wug task:

Berko, Jean. "The child’s learning of English morphology." Word 14, no. 2-3 (1958): 150-177.

The natural class theory of segmental distances:

- Frisch, Stefan. Similarity and frequency in phonology. Northwestern University, 1996.
- Frisch, Stefan A., Janet B. Pierrehumbert, and Michael B. Broe. Similarity avoidance and the OCP. Natural language & linguistic theory 22, no. 1 (2004): 179-228.

Word distances based on segmental distances:

- Albright, Adam, and Bruce Hayes. Rules vs. analogy in English past tenses: A computational/experimental study. Cognition 90, no. 2 (2003): 119-161.
- Dawdy-Hesterberg, Lisa Garnand, and Janet Breckenridge Pierrehumbert. Learnability and generalisation of Arabic broken plural nouns. Language, cognition and neuroscience 29, no. 10 (2014): 1268-1282.

Features:

- Siptár, Péter, and Miklós Törkenczy. The phonology of Hungarian. OUP Oxford, 2000.

Training data:

- Rácz, Péter, and Ágnes Lukács. Morphological convergence and sociolinguistic salience: an experimental study. (2023) [https://osf.io/preprints/psyarxiv/zqwxv](https://osf.io/preprints/psyarxiv/zqwxv).
- Nemeskey, Dávid Márk (2020). Natural Language Processing methods for Language Modeling. PhD thesis. Eötvös Loránd University. [https://hlt.bme.hu/en/resources/webcorpus2](https://hlt.bme.hu/en/resources/webcorpus2)

GCM, KNN:

- Nosofsky, Robert M. "The generalized context model: An exemplar model of classification." Formal approaches in categorization (2011): 18-39.
- Peterson, Leif E. "K-nearest neighbor." Scholarpedia 4, no. 2 (2009): 1883.

MGL:

- Albright, Adam, and Bruce Hayes. "Modeling English past tense intuitions with minimal generalization." In Proceedings of the ACL-02 workshop on Morphological and phonological learning, pp. 58-69. 2002.
- Mikheev, Andrei. "Automatic rule induction for unknown-word guessing." Computational Linguistics 23, no. 3 (1997): 405-423.

The three variable patterns:

- 'cselekszenek/cselekednek': Rácz, Péter, Péter Rebrus, and Miklós Törkenczy. "Attractors of variation in Hungarian inflectional morphology." Corpus Linguistics and Linguistic Theory 17, no. 2 (2021): 287-317.
- 'hotelban/hotelben': Hayes, Bruce, Péter Siptár, Kie Zuraw, and Zsuzsa Londe. "Natural and unnatural constraints in Hungarian vowel harmony." Language (2009): 822-863.
- 'lakok/lakom': Rácz, Péter. "Frequency and prototypicality determine variation in the Hungarian verbal 1 SG. INDEF." Acta Linguistica Academica. An International Journal of Linguistics (Until 2016 Acta Linguistica Hungarica) 66, no. 4 (2019): 601-620.

## Appendix

Predictions with test word labels:

```{r viz2, fig.width = 10, fig.height = 54}
plots = tests %>% 
  mutate(base2 = str_extract(base, '^[^ ]+(?=( |$))')) %>%
  select(variation,base2,log_odds,tof,knn,gcm) %>% 
  pivot_longer(cols = c(tof,knn,gcm), names_to = 'model', values_to = 'weight') %>% 
  group_by(variation,model) %>% 
  nest() %>% 
  mutate(
    title = glue('{variation}: best {model}'),
    plot = map2(title, data, ~ 
           {
             .y %>% 
               ggplot(aes(x = weight, y = log_odds, label = base2)) +
                  ggrepel::geom_label_repel(max.overlaps = 50) +
                  theme_bw() +
                  geom_smooth(method = mgcv::gam) +
                  ggtitle(.x)
               }
           )
  )

# plots from list
wrap_plots(plots$plot, ncol = 1)
```
